---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading applicable libraries
```{r loadLibraries}
   library(plyr)
   library(dplyr)
   library(knitr)
   library(ggplot2)
```

## set Knitr global options to always echo code chunks
```{r globaloptions}
   knitr::opts_chunk$set(echo = TRUE) 
```

## Loading and preprocessing the data
Show any code that is needed to

1. Load the data (i.e. read.csv())

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r loaddata}
   activitydf <- read.csv("activity.csv", header=TRUE, sep=",", colClasses=c("numeric","character","numeric"), stringsAsFactors = FALSE, na.strings = "NA")
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day

```{r totalnumberofstepsperday}
   mydf <- activitydf[complete.cases(activitydf),]
   steps_per_day <- aggregate(steps ~ date, data = mydf, FUN = "sum", na.rm = TRUE)
   steps_per_day
```

2. If you do not understand the difference between a histogram and a barplot, research the difference between them.

Make a histogram of the total number of steps taken each day

```{r meanstepsperday}
   hist(steps_per_day$steps, main = "Total Steps Per Day", xlab="Number of Steps")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r meanandmedian}
   mean(steps_per_day$steps)
   median(steps_per_day$steps)
```
** The mean number of total steps taken per day was 10766

** The median number of total steps taken per day was 10765

## What is the average daily activity pattern?
Processing the data per time interval

```{r averagetimeinterval}
   average_daily_activity <- aggregate(steps ~ interval, data = activitydf, FUN = "mean", na.rm = TRUE)
   head(average_daily_activity)
```

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r timeseriesplot}
   ggplot(average_daily_activity, aes(x =interval , y=steps)) +
   geom_line(color="red", size=1) +
   labs(title = "Average Daily Steps", x = "Interval", y = "Average number of Steps Per Day")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r intervalmaxsteps}
   average_daily_activity[which.max(average_daily_activity$steps),]
```

** Interval 835 has the maximum of 206.17 steps

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r nainstances}
   sum_missing_values <- sum(is.na(activitydf$steps))
   sum_missing_values
```

** The total sumber of NA occurences in steps column is 2304 

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc. 

I chose the mean for the 5-minute interval.

```{r intervalmean}
   interval_average <- tapply(activitydf$steps, activitydf$interval, mean, na.rm=TRUE, simplify = TRUE)
   interval_average
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r fillingnas}
   na_filled_df <- activitydf
   na_logical <- is.na(na_filled_df$steps)
   na_filled_df$steps[na_logical] <- interval_average[as.character(na_filled_df$interval[na_logical])]
   ## head of the original dataset including NA
   ## head(activitydf)
   ## sum(is.na(activitydf))
   ## head of the simulated dataset has no NA
   head(na_filled_df)
   sum(is.na(na_filled_df))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r numberofsactvstepsperday}
   sum_activity_filled_df <- aggregate(steps ~ date, data = na_filled_df, FUN = "sum", na.rm = TRUE)
   hist(sum_activity_filled_df$steps, main = "Total Steps Per Day", xlab="Number of Steps")
   sum_activity_filled_df
```

```{r computemeanafterfill}
   steps_postfil_mean <- mean(sum_activity_filled_df$steps, na.rm=TRUE)
   steps_postfil_mean
   steps_postfil_median <- median(sum_activity_filled_df$steps, na.rm=TRUE)
   steps_postfil_median
```

Do these values differ from the estimates from the first part of the assignment? 

Yes there is a slight difference with the median value.

** The mean number of total steps taken per day was 10766

** The median number of total steps taken per day was 10766

What is the impact of imputing missing data on the estimates of the total daily number of steps?

** Both the mean & median have become the same, yet the median has changed slightly from the previous results: 10765 --> 10766. 

The number of occurences (a.k.a frequency) has increased as expected

## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r createwktype}
   na_filled_df$date <- as.POSIXct(na_filled_df$date)
   na_filled_df <- mutate(na_filled_df, weektype = ifelse(weekdays(na_filled_df$date)=="Saturday" | weekdays(na_filled_df$date)=="Sunday", "Weekend", "Weekday"))
   head(na_filled_df)
```

```{r sortwktypedataset}
   # group dataset by interval & weektype then compute mean on steps
   na_filled_df2<- na_filled_df %>% group_by(interval, weektype) %>%
        summarise(steps = mean(steps, na.rm=TRUE))
head(na_filled_df2)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r seriesplotweektype}
p <- ggplot(na_filled_df2, aes(x =interval , y=steps, color=weektype)) +
       geom_line() +
       labs(title = "Average Daily Steps by WeekType", x = "Interval", y = "Number of Steps") +
       facet_wrap(~weektype, ncol = 1, nrow=2)
p
```

** The subject appears to be more active earlier in the day during the weekday and is more active throughout the weekend. 

** This is most likely due to the subject sitting a good portion of the time and hence explain the lesser movement during the weekdays' working hours.

